The hidden manifold distance for functional data

### 2019-06-26 index.Rmd
prepping compare_methods

---
title: "The hidden manifold distance for functional data"
output:
  html_document:
    df_print: paged
    number_sections: yes
  pdf_document: default
---

@@

# Introduction

@@

## Setting
Describe preprocessing of discretely observed functional data: In the classical approach to analysing such data, we first obtain a functional version $\tilde X_1,\ldots,\tilde X_n$ of the raw data (by either smoothing or using PACE), next perform FPCA to reduce the dimension of the data, and then proceed by employing multivariate techniques.

@@

# Preliminaries

@@

## Basic differential geometry
This is take from a paper on [metric learning](http://proceedings.mlr.press/v32/linb14.pdf).
Let $(\M,g)$ be a $d$-dimensional Riemannian manifold, where $g$ is a Riemannian metric tensor on $\M$. This metric tensor can be used to assign a metric on the manifold. For each point $p$ on the manifold, the Riemannian metric tensor $g$ has an inner product $g_p$ on the tangent space $T_p \M$. The norm of a tangent vector $v \in T_p \M$ is defined as $$||v|| = \sqrt{g_p(v,v)}$$
If $\gamma: [a,b] \subset \R \to \M$ is a smooth curve, its length is then defined as
$$ l(\gamma) := \int_a^b || \frac{\,d\gamma}{\,d t}(t) || \,dt $$
The geodesic distance between two points $p,q$ on the manifold $\M$, based on this metric tensor $g$, is defined as
$$ d_g(p,q):=\inf \{l(\gamma): \gamma:[a,b] \to \M \text{ piecewise smooth}, \gamma(a) = p, \gamma(b) = q]  \}.$$

## IsoGeo

@@

# Proposed method for estimating geodesic distances

@@

## Selection of tuning parameters

@@

## Alternative estimators of geodesic distance

@@

## Performance metrics
We describe three different metrics we use to assess the quality of a pairwise geodesic distance estimator. This is the ROC curve with $\epsilon$ on the $x$-axis and degree to which near-$\epsilon$ isometry holds, i.e.\ the percentage of estimated pairwise distances between $1-\epsilon$ and $1+\epsilon$ of the truth pairwise distance.

## Simulation scenarios
Acknowledge that we are sampling on a very concentrated measure to make sure functional manifold is well sampled. Cite Diaconis et. al's "Sampling from a manifold". How to sample properly from a functional manifold could be interesting future work.

@@

## The cost of employing proposed method when manifold is flat

@@

# Distance-based functional classification

@@

For simplicity, assume the task is bianry classification. Associated to each functional object $x$ is a binary $y$ indicating class membership.
Consider the classifier proposed in Ferraty and Vieu (2003,2006) which is a functional version of the Nadaraya-Watson kernel estimator of class membership probabilities:
$$
\hat p(y = 0 | x) \frac{ \sum_{i=1}^n K[h^{-1} d(x,x_i)] 1(y_i = 0) }{ \sum_{i=1}^n K[h^{-1} d(x,x_i)] }
$$
We shall compare our method to using $L_2$ distance, possibly weighted, and with curve registration already accomplished.

@@

Datasets used in functional manifold papers
* Berkeley growth, yeast cell cycle gene expression (can't find this publicly) in Chen and Muller
* Tecator in Lin and Yao's contamination paper
* Berkeley growth, gait cycle in Jean-Michel Loubes' robust isomap paper

>>>>

---
title: "The hidden manifold distance for functional data"
output:
  pdf_document: 
    number_sections: yes
  html_document:
    df_print: paged
    number_sections: yes
---

@@

# Introduction [Susan]

@@

## Setting [Susan]
Describe preprocessing of discretely observed functional data: In the classical approach to analysing such data, we first obtain a functional version $\tilde X_1,\ldots,\tilde X_n$ of the raw data (by either smoothing or using PACE), next perform FPCA to reduce the dimension of the data, and then proceed by employing multivariate techniques.

@@

# Preliminaries [Susan]

@@

## Basic differential geometry [Susan]
This is take from a paper on [metric learning](http://proceedings.mlr.press/v32/linb14.pdf).
Let $(\M,g)$ be a $d$-dimensional Riemannian manifold, where $g$ is a Riemannian metric tensor on $\M$. This metric tensor can be used to assign a metric on the manifold. For each point $p$ on the manifold, the Riemannian metric tensor $g$ has an inner product $g_p$ on the tangent space $T_p \M$. The norm of a tangent vector $v \in T_p \M$ is defined as $$||v|| = \sqrt{g_p(v,v)}$$
If $\gamma: [a,b] \subset \R \to \M$ is a smooth curve, its length is then defined as
$$ l(\gamma) := \int_a^b || \frac{\,d\gamma}{\,d t}(t) || \,dt $$
The geodesic distance between two points $p,q$ on the manifold $\M$, based on this metric tensor $g$, is defined as
$$ d_g(p,q):=\inf \{l(\gamma): \gamma:[a,b] \to \M \text{ piecewise smooth}, \gamma(a) = p, \gamma(b) = q]  \}.$$

## IsoGeo [Marie]

@@

# Proposed method for estimating geodesic distances [Marie]

@@

## Selection of tuning parameters [Marie]
Describe heuristics.

@@

## Alternative estimators of geodesic distance [Marie]

@@

## Performance metrics [Susan]
We describe three different metrics we use to assess the quality of a pairwise geodesic distance estimator. This is the ROC curve with $\epsilon$ on the $x$-axis and degree to which near-$\epsilon$ isometry holds, i.e.\ the percentage of estimated pairwise distances between $1-\epsilon$ and $1+\epsilon$ of the truth pairwise distance.

## Simulation scenarios [Susan]
Acknowledge that we are sampling on a very concentrated measure to make sure functional manifold is well sampled. Cite Diaconis et. al's "Sampling from a manifold". How to sample properly from a functional manifold could be interesting future work.

@@

## The cost of employing proposed method when manifold is flat [Susan/Marie?]

@@

# Distance-based functional classification [Marie]

@@

For simplicity, assume the task is bianry classification. Associated to each functional object $x$ is a binary $y$ indicating class membership.
Consider the classifier proposed in Ferraty and Vieu (2003,2006) which is a functional version of the Nadaraya-Watson kernel estimator of class membership probabilities:
$$
\hat p(y = 0 | x) \frac{ \sum_{i=1}^n K[h^{-1} d(x,x_i)] 1(y_i = 0) }{ \sum_{i=1}^n K[h^{-1} d(x,x_i)] }
$$
We shall compare our method to using $L_2$ distance, possibly weighted, and with curve registration already accomplished. Describe alternative methods in detail.

The bandwidth in the classifier should be tuned individually for each method. Also we might need to tune MDS dimension $s$ since in real data, the dimension of the manifold might be much higher than encountered in the simulation scenarios where it never goes above 2.

@@

Datasets used in functional manifold papers

* Berkeley growth, yeast cell cycle gene expression (can't find this publicly) in Chen and Muller
* Tecator in Lin and Yao's contamination paper
* Berkeley growth, gait cycle in Jean-Michel Loubes' robust isomap paper
